\section{Introduction}
\label{sec:intro}

Adaptive radiotherapy, longitudinal disease monitoring, and remote-sensing
change detection all share a structural prior: a prior-session observation
of the same scene or patient is available at inference time.  Existing
segmentation pipelines either discard this prior, concatenate it as an
input channel, or fuse it with shallow siamese features.  We argue that a
more principled approach is to treat the prior session as a \emph{memory bank}
of tokens attended to by a frozen foundation backbone, and to simultaneously
infer a latent code for the patient's \emph{treatment response phenotype} —
a compact representation of how their tumour changes under therapy.

CSM-SAM instantiates this idea on top of SAM2~\cite{ravi2024sam2}.
The cross-session memory attention bridges week-0 (pre-RT) and week-2/3
(mid-RT) scans without registration.  A variational change latent
$\mathbf{z}$ is inferred from paired pre/mid features at training time and
from prior + cross-patient retrieval at inference, then FiLM-conditions the
attention query so that the retrieval pattern depends on the inferred phenotype.

Our central claim is that $\mathbf{z}$, trained \emph{purely on the
segmentation objective with no phenotype labels}, nonetheless
organises into clusters that correlate with the volume-change ratio $\Delta V$
— a standard clinical indicator of treatment response.  This is an
\emph{emergent} property of the representation, not a supervised output.
It is the fact that makes this paper distinguishable from the body of
memory-augmented or change-detection literature: the model discovers
radiobiological structure without being asked.

\paragraph{Contributions.}
\begin{enumerate}
\item \textbf{Cross-session memory attention on SAM2.}  We extend
  SAM2's within-session memory mechanism to the inter-session
  (pre-RT $\to$ mid-RT) setting, conditioned on continuous elapsed-time
  embeddings.  The module applies unchanged to six diverse benchmarks
  (medical longitudinal, binary CD, semantic CD) without per-dataset
  re-engineering.

\item \textbf{Variational change latent with emergent phenotype structure.}
  A posterior encoder $q(\mathbf{z}\mid\mathbf{f}_{\text{pre}},
  \mathbf{f}_{\text{mid}}, t)$ and learned prior $p(\mathbf{z}\mid
  \mathbf{f}_{\text{pre}}, t)$ are trained end-to-end via a KL-annealed
  segmentation objective.  At evaluation we show that $\boldsymbol{\mu}_q$
  clusters by treatment response phenotype (Spearman $\rho > 0.5$ with
  $\Delta V$, $p < 0.01$) without any phenotype supervision.

\item \textbf{Retrieval-based posterior inference.}  At test time the
  prior is blended with a weighted mean of stored $\mathbf{z}_i$ from
  the $K$ nearest training patients, turning cross-patient retrieval into
  approximate posterior inference over the change latent.

\item \textbf{State-of-the-art on HNTS-MRG 2024 Task~2.}  We exceed the
  challenge winner's aggDSC of 0.733 while using fewer trainable parameters
  than any top-5 entry.
\end{enumerate}
