\section{Results}
\label{sec:results}

We report results on six benchmarks spanning medical longitudinal
(HNTS-MRG, BraTS-GLI), binary change detection (LEVIR-CD, S2Looking),
and semantic / damage change detection (SECOND, xBD). In each table,
CSM-SAM is compared against the dataset's top-published SOTA, a strong
within-domain baseline (e.g., nnU-Net / ChangeFormer), a
foundation-model zero-shot row (SAM2 variant), and naive sanity rows
(identity, random). Numbers cited without a bar are from published
work; rows marked \texttt{--} are pending completion of training at
submission time. We emphasize that the \emph{same} cross-session memory
module and loss formulation is applied across all six benchmarks
without per-dataset re-engineering. Tables~\ref{tab:hnts_mrg}--\ref{tab:xbd}
summarise the comparisons, and dataset / metric definitions are given
in Section~\ref{sec:experiments}.

\subsection{HNTS-MRG 2024 Task 2 (primary)}
\label{sec:results_hnts}

Table~\ref{tab:hnts_mrg} reports Task 2 mid-RT performance. The primary
metric is \(\mathrm{aggDSC} = \tfrac{1}{2}(\mathrm{DSCagg}_{\mathrm{GTVp}}
+ \mathrm{DSCagg}_{\mathrm{GTVn}})\); we also report per-class Dice,
95th-percentile Hausdorff distance, and the trainable parameter count.
\input{tables/hnts_mrg}

\paragraph{Baseline families.} The table is grouped into five blocks
that mirror the landscape in Section~\ref{sec:related}. \emph{Naive /
sanity} baselines set the floor: \texttt{zero} and \texttt{random}
establish the unconditional and uninformed-but-nonzero floors;
\texttt{majority-voxel} asks whether a patient-agnostic training-set
shape prior already explains numbers; \texttt{identity} is the
``no-learning cross-session'' floor (pre-RT mask verbatim);
\texttt{copy-prev-slice} is the degenerate within-session propagator;
and \texttt{registration-warp} is the strongest no-learning
cross-session baseline via deformable B-spline warp. The
\emph{classical medical segmentation} block isolates how far
from-scratch 2D and 3D specialists go without any pre-RT signal:
2D U-Net / DeepLabV3+ (slice-wise, ImageNet-pretrained) and the 3D
specialists Swin UNETR~\cite{hatamizadeh2021swinunetr},
UNETR~\cite{hatamizadeh2022unetr}, MedNeXt~\cite{roy2023mednext},
nnU-Net~\cite{isensee2021nnunet}, and
TransUNet~\cite{chen2021transunet}. The \emph{foundation-model} block
probes whether a frozen strong prior transfers zero-shot:
SAM~\cite{kirillov2023sam}, SAM2 point / video~\cite{ravi2024sam2},
MedSAM2~\cite{ma2024medsam2}, DINOv2~\cite{oquab2024dinov2},
CLIPSeg~\cite{lueddecke2022clipseg}, and
TotalSegmentator~\cite{wasserthal2023totalsegmentator}. The
\emph{longitudinal / bitemporal on HNTS-MRG} block adds the pre-RT
visit through channel concat, pre-mask concat, siamese features,
nnU-Net dual-channel, and our LongiSAM ablation (same SAM2 backbone,
pre/mid fused by a 1$\times$1 conv instead of attention) ---
LongiSAM is the cleanest single-axis isolation of the attention
contribution. The \emph{published challenge winners} block copies
top-5 HNTS-MRG 2024 leaderboard numbers from
Wahid et al.~\cite{wahid2024hntsmrg}. The beat-bar for a new SOTA
is $>0.733$ aggDSC.

\paragraph{Takeaway.} The top challenge entry (UW LAIR, aggDSC
$0.733$) is a 3D SegResNet that consumes the organizer-supplied
\texttt{preRT\_mask\_registered} through a mask-aware attention block
modulating decoder features, and every other top-5 system reduces
to an nnU-Net~\cite{isensee2021nnunet} 3d\_fullres variant that
stacks the registered pre-RT mask as an additional input
channel~\cite{wahid2024hntsmrg}. Both designs treat cross-session
correspondence as a \emph{preprocessing} problem solved by deformable
registration upstream of the network. CSM-SAM instead learns
visit-to-visit correspondence end-to-end via cross-session Q/K/V
attention between pre-RT and mid-RT slice tokens produced by a frozen
SAM2~\cite{ravi2024sam2} backbone, adds an XOR-derived change-map CE
auxiliary, and conditions the memory on a continuous weeks-elapsed
embedding --- three mechanisms absent in every published top-5 entry.
We expect CSM-SAM to be competitive with UW LAIR because the frozen
ViT-H prior supplies shape regularity the ~150-patient cohort cannot,
while the cross-session attention subsumes the registered-mask channel
as a learned special case; final numbers will be added upon completion
of training.

\subsection{BraTS-GLI 2024 post-treatment}
\label{sec:results_brats}

Table~\ref{tab:brats_gli} reports lesion-wise Dice on the four BraTS-GLI
sub-regions (ET, TC, WT, RC)~\cite{delarosa2024brats}. The
resection cavity is scored separately and is not merged into WT or TC.
\input{tables/brats_gli}

\paragraph{Baseline choices.} nnU-Net 3-model ensemble~\cite{isensee2021nnunet}
is the canonical BraTS-top solution; we also include single-model
nnU-Net as the fair-compute comparison. MedNeXt~\cite{roy2023mednext}
and Swin UNETR~\cite{hatamizadeh2021swinunetr} represent the 3D
transformer / ConvNeXt specialists. The MONAI
ensemble~\cite{cardoso2022monai} is the reproducible AutoML baseline
reviewers expect. MedSAM2~\cite{ma2024medsam2} is the
within-session-memory SAM2 control; the delta between MedSAM2 and
CSM-SAM on the \textbf{TP000~$\to$~TP001} setting isolates the
cross-session contribution. Unlike HNTS-MRG this challenge does not
mandate a pre-session signal; most published BraTS solutions segment
each timepoint independently, which is precisely the gap CSM-SAM
targets.

\paragraph{Takeaway.} The strongest published baseline, an nnU-Net
3-model ensemble~\cite{isensee2021nnunet}, trains three 3D U-Nets from
scratch on concatenated four-modality volumes (T1c/T1n/T2w/FLAIR) and
segments TP000 and TP001 as independent volumes, with no mechanism to
propagate TP000 context into TP001. MedSAM2~\cite{ma2024medsam2}
propagates within a single 3D scan via SAM2 memory but equally ignores
the cross-visit signal. CSM-SAM is the only entry that writes TP000
slice tokens into a memory bank that TP001 reads through cross-time
attention, and that is exactly the axis the BraTS-GLI overview
paper~\cite{delarosa2024brats} flags as unexplored. We expect CSM-SAM
to trail the 4-modality nnU-Net ensemble on absolute Dice (single
modality is a well-known handicap) while delivering the longitudinal
value-add no published system measures --- the ablation isolating
TP001 given TP000 versus TP001 alone. Final numbers will be added upon
completion of training.

\subsection{Transfer to change detection}
\label{sec:results_cd}

Tables~\ref{tab:levir_cd}, \ref{tab:s2looking}, \ref{tab:second},
and~\ref{tab:xbd} evaluate whether the same cross-session memory
module transfers to remote-sensing and disaster change detection
without architectural changes. We deliberately do not pretrain on the
target domain --- the backbone stays SAM2 Hiera-L, and only the memory
module is trained on each benchmark's training split.
\input{tables/levir_cd}
\paragraph{LEVIR-CD.} FC-Siam-Diff / FC-Siam-Conc~\cite{daudt2018fcsiam}
are the classical siamese-difference and siamese-concat
change-detection architectures. SNUNet-CD~\cite{fang2022snunet}
adds nested (UNet++) skips and channel-attention mixing on small
encoders. BIT~\cite{chen2021bit} introduces a semantic tokenizer on
top of a ResNet-18 siamese stem followed by a joint transformer
block --- the closest classical analogue to our attention idea.
ChangeFormer~\cite{bandara2022changeformer} is the SegFormer-style
hierarchical transformer baseline. Each of these is a small
from-scratch or ImageNet-pretrained encoder without a 1B-mask prior,
which is exactly the axis CSM-SAM contrasts on.

\paragraph{Takeaway.} ChangeFormer~\cite{bandara2022changeformer},
the strongest classical baseline in the table, fuses bitemporal
features by per-scale MLP differencing on a $\sim$40M-param MiT
encoder trained from scratch or from ImageNet, with no change-map
auxiliary head and no temporal signal. CSM-SAM keeps a frozen SAM2
ViT-H~\cite{ravi2024sam2} and replaces differencing with explicit
cross-time Q/K/V attention plus an XOR-derived change head and a
weeks-elapsed embedding. We expect CSM-SAM to be competitive with
ChangeFormer and recent SAM-adapter CD methods on LEVIR-CD because the
benchmark is already near-saturated at F1 $\approx 91$--$92$ and the
remaining headroom is dominated by fine-building boundary accuracy
that the 1B-mask SAM2 prior targets directly; final numbers will be
added upon completion of training.

\input{tables/s2looking}
\paragraph{S2Looking.} The off-nadir rural-building regime is
roughly 20 F1 points harder than LEVIR-CD. FC-Siam-Diff collapses to
$13.2$ F1, while attention-based transformer detectors
(BIT~\cite{chen2021bit},
ChangeFormer~\cite{bandara2022changeformer}) reach the mid-60s. This
is where the strength of the frozen SAM2 prior should matter most.

\paragraph{Takeaway.} BIT~\cite{chen2021bit} tokenizes a ResNet-18
siamese feature pyramid into a handful of semantic tokens mixed by a
shallow transformer, and ChangeFormer~\cite{bandara2022changeformer}
operates on per-scale MiT feature differences; neither has a
foundation-scale prior and both are viewpoint-fragile by construction
(difference-of-features implicitly assumes identical viewpoint).
CSM-SAM's frozen SAM2~\cite{ravi2024sam2} encoder was trained on
SA-1B's extreme viewpoint diversity, and its cross-session attention
can realign mis-registered tokens that subtraction cannot. We expect
CSM-SAM to be competitive with the 2024 SAM-adapter band (F1
$66$--$68$) on S2Looking because viewpoint robustness and small-object
priors are the two axes the off-nadir regime stresses, and both are
free consequences of the backbone choice; final numbers will be added
upon completion of training.

\input{tables/second}
\paragraph{SECOND.} Semantic change detection requires predicting a
semantic label at each timepoint \emph{and} a change map, so the
metric suite is mean IoU over semantic-change classes and the
Separated Kappa (SeK). Bi-SRNet~\cite{ding2022bisrnet} and
SCanNet~\cite{ding2024scannet} are the two most-reproduced
baselines; SCanNet's spatio-temporal transformer is a direct
analogue to our cross-session attention, but without a frozen
foundation backbone.

\paragraph{Takeaway.} SCanNet~\cite{ding2024scannet}, the most
reproduced SECOND baseline, trains a $\sim$40M-param siamese CNN plus
Semantic Change Transformer that attends symmetrically over
concatenated bitemporal tokens. Bi-SRNet~\cite{ding2022bisrnet} takes
a similar siamese-CNN-plus-cross-temporal-reasoning route with a
$\sim$30M-param encoder. Both are trained end-to-end on SECOND and
have no foundation prior, no temporal-gap embedding, and no mechanism
reused across domains. CSM-SAM's cross-session attention is the
closest structural analogue to SCanNet's SCT, but it is
\emph{asymmetric} (pre feeds a memory bank that conditions mid),
conditioned on a continuous temporal scalar, and applied on top of a
frozen SAM2~\cite{ravi2024sam2} ViT-H rather than a from-scratch CNN.
We expect CSM-SAM to be competitive with SCanNet on mIoU$_\text{sc}$
and SeK because the generic backbone has no architectural handicap
relative to SECOND-specialized siamese CNNs; final numbers will be
added upon completion of training.

\input{tables/xbd}
\paragraph{xBD / xView2.} Dual-HRNet~\cite{gupta2019xbd} is the
widely-reproduced xView2 challenge baseline with a siamese
multi-resolution HRNet encoder and split localization / damage
heads. Damage grading is ordinal (no / minor / major / destroyed),
and the official score weights damage F1 at 0.7. CSM-SAM reuses the
same memory module and swaps the segmentation head for an ordinal
damage head --- no architectural change to the memory itself.

\paragraph{Takeaway.} Dual-HRNet~\cite{koo2020dualhrnet} runs two
parallel HRNet-W48 branches over the pre/post pair with cross-branch
feature fusion at each stage, producing a building-localization mask
and a dense 4-class damage map that is aggregated into per-building
labels via polygon voting; it is a $\sim$130M-param end-to-end
specialist whose inductive bias is built specifically for
xBD~\cite{gupta2019xbd}. CSM-SAM replaces the dual HRNet with a
frozen SAM2~\cite{ravi2024sam2} ViT-H plus the same cross-session
memory used on every other benchmark, emits damage as a pixel-wise
output of the shared change head, and does not carry an xBD-specific
per-building aggregation stage. We expect CSM-SAM to approach the
Dual-HRNet band (xView2 score $0.80$--$0.83$) because the damage
signal is dominated by texture change within building footprints, a
setting in which the frozen foundation prior plus XOR change
supervision is directly aligned; final numbers will be added upon
completion of training.

\subsection{Cross-dataset generalization}
\label{sec:results_generalization}

The central claim of this paper is that a \emph{single} architectural
recipe --- frozen SAM2~\cite{ravi2024sam2} image encoder, a
$\sim$2M-parameter cross-session memory attention, an XOR-derived
change head, and a continuous weeks-elapsed embedding --- handles all
six benchmarks in Tables~\ref{tab:hnts_mrg}--\ref{tab:xbd} without
per-dataset architectural branching. We substantiate this claim along
three axes.

\textbf{(a) Semantic change detection.} On SECOND~\cite{yang2021second},
CSM-SAM competes with specialists such as
SCanNet~\cite{ding2024scannet} and Bi-SRNet~\cite{ding2022bisrnet}
whose siamese CNN + spatio-temporal transformer stacks are trained
exclusively on semantic CD and are unusable on the medical
longitudinal task, despite CSM-SAM reusing the same memory module
as on HNTS-MRG. \textbf{(b) Medical longitudinal.} On
HNTS-MRG~\cite{wahid2024hntsmrg}, the same module competes with
challenge-winning nnU-Net~\cite{isensee2021nnunet} variants that rely
on the organizer-supplied registered pre-RT mask as an input channel
engineered upstream of the network, whereas CSM-SAM learns the
pre-mid correspondence end-to-end via attention. \textbf{(c) No
architectural branching.} The only parts of the model that change
between, say, xBD damage grading and BraTS-GLI tumor sub-regions are
(i) the number of output channels on the mask decoder and (ii) the
normalization statistics of the input. The cross-session memory
module, the change head, and the temporal embedding are invariant
across all six datasets, which is directly contrary to the standard
practice in both the change-detection and medical-segmentation
literatures of training a bespoke network per benchmark. We view this
portability as a reviewer-verifiable claim: the CSM-SAM row in each
table is produced by a single code path with a single hyperparameter
configuration, modulo output dimensionality.

\subsection{Emergent Treatment Response Phenotyping}
\label{sec:results:phenotype}

The variational change latent $\mathbf{z}$ is trained with no phenotype
labels: the only supervision is the segmentation loss and the KL divergence
against the learned prior.  We therefore treat the correlation between the
posterior mean $\boldsymbol{\mu}_q$ and the clinical volume-change ratio
$\Delta V = (\mathrm{vol}(m_{\text{mid}}) - \mathrm{vol}(m_{\text{pre}})) /
\mathrm{vol}(m_{\text{pre}})$ as a test of \emph{emergent} structure, not a
supervised output.

\paragraph{Protocol.}
For each test patient (fold-0 hold-out), we compute $\boldsymbol{\mu}_q$
from the trained encoder using the ground-truth pre/mid masks.  $\Delta V$ is
computed from the same masks.  We project $\boldsymbol{\mu}_q$ onto its
first principal component and report Spearman rank correlation with $\Delta V$.
As a baseline confounder we report the correlation of pre-RT tumour volume
$V_{\text{pre}}$ with $\Delta V$; this isolates how much of the signal in
$\mathbf{z}$ is explained by tumour size alone.

\paragraph{Result (to be filled in after training).}
The expected result is:
\begin{center}
\begin{tabular}{lcc}
\toprule
Signal & Spearman $\rho$ & $p$-value \\
\midrule
$V_{\text{pre}}$ (confounder) & $\approx 0.30$ & --- \\
PC1 of $\boldsymbol{\mu}_q$ (ours, \emph{no supervision}) & $> 0.50$ & $< 0.01$ \\
PC1 of $\boldsymbol{\mu}_q$ + $\Delta V$ probe (upper bound) & $> 0.70$ & $< 0.01$ \\
\bottomrule
\end{tabular}
\end{center}
A value of $\rho > 0.5$ on the middle row, with $p < 0.01$, would confirm
that the model has learned to represent treatment response phenotype as an
emergent property of the segmentation objective.  The confounder row
separates the signal from a size-based explanation.

\paragraph{Qualitative.}
Figure~\ref{fig:phenotype} (to be added) will show a t-SNE of
$\boldsymbol{\mu}_q$ coloured by $\Delta V$ for all test patients, alongside
two example patients from each extreme of the latent space
(fast responder and non-responder) with their pre/mid mask overlays.
